## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

if (logical statement) { do something } else { do something else }

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10
if (a<=5) {
  print(a)
} else {
  print("it's more than 5")
}
#R uses braces {} to figure out where code begins & ends
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 25
b <- 23
if (a>b) {
  print ("a is greater than b")
} else if (a == b){
  print ("a and b are the same")
} else {
  print("a must be less than b")
}

```

Neat! We can also nest conditionals just like we do in Python.

```{r}
c <- 445
d <- 256
e <- 622
if (e + c > d + e | d + c < 500) {
  print("the first test is true")
if (d + c > 500){
  print("tests 1 and 2 are true")
}
} else {
  print("none of these are true")
}
    
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
my_vector <- c(5, 3, 7,9,1,23, 6,2, 7, 55)
for (i in my_vector){
  i = i * 2
  print(i)
}

```

```{r}
for (value in 1:5) {
  #Paste is a concatenation function
  print(paste(value, "is greater than", value -1))
} #don't need to add spaces
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1)
#change first, third and and sixth elements to value 6
for (i in c(1,3,6)){
  a_vector[i] <- 6
}
a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow =2, ncol=2)

for (row in 1:2){
  for (col in 1:2){
    mat[row,col] <- 0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0
while (b < 10){
  #increase by 2
  b <-  b+2
  print (b)
}
```

## Functions

When should we use functions?

```{r}
#use when doing something 2-3 times or if testing to save memory
our_function <- function(optional_params){
  # do some stuff
}
our_function(#pass something in)
```

An example:

```{r}
#start w a vector of temperatures
test_temp <- c(16.550540, 26.942005, 32.591088, 8.580016, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735)

#then we want a function that converts from C to F
c_to_f <- function(dat){
  temp_farh <- dat*(9/5) + 32
  
  return(temp_farh)
}
#R doesn't care about the data passed in
print(c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
#mapping over the array without a for loop
data_fahr <- c_to_f(test_temp)
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow <- 2)
#1 is for rows, 2 is for columns
apply(test_mat, 2, sum) 
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix( c(16.550540, 26.942005, 32.591088, 8.580016, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735), nrow =2)
apply(test_temp_mat, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    #1. Create a function with two arguments that will return an array of the first n multiples of x.
    count_by <- function(x, n){
      ct_array <- c(1:n)
      for (i in ct_array){
        ct_array[i] <- x*i
      }
      print(ct_array)
    }

    count_by(2, 5)
    #top answer was:
    #count_by <- function(x, n){
    #  seq(from=x,by=x,length=n)
    #}
    ```

    ```{r}
    #2. Create a function which translates a given DNA string into RNA. (replace T with U in string)
    dna_to_rna <- function(dna){
        gsub("T","U",dna)
      }

    dna_to_rna("TTTT")
    #same as top answer :)
    ```

    ```{r}
    #3 Given a list of unique numbers sorted in ascending order, return a new list so that the values increment by 1 for each index from the minimum value up to the maximum value (both included).
    pipe_fix <- function(nums){

      my_seq <- seq(nums[1], tail(nums,n=1), by=1)
      my_seq
    }
    pipe_fix(c(6, 9))
    #top answer (was considering min, max but wasn't sure it worked in R)
    #pipe_fix <- function(nums){
    #  return(min(nums):max(nums))
    #}
    ```

    ```{r}
    #4 Complete the method/function so that it converts dash/underscore delimited words into camel casing. The first word within the output should be capitalized only if the original word was capitalized (known as Upper Camel Case, also often referred to as Pascal case). The next words should be always capitalized.
    to_camel_case <- function(text){
      gsub("([[:punct:]])(\\w)","\\U\\2", text, perl=TRUE)
    }
    to_camel_case("mixed-separator_example")
    #top answer
    #to_camel_case <- function(text){
    #  gsub('[-_](.)','\\U\\1',text,perl=T)
    #}
    ```

    ```{r}
    #5 Remove all values from list a, which are present in list b keeping their order.
    array_diff = function(a, b) {
      a[! a %in% b]
    }
    array_diff(c(1, 2, 2), 1)
    #very similar to top answer
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    #bonus points for mapping over an array or could generate a vector

    check_multiple <- function(i){
      if ((i %% 3 == 0) & (i %% 5 == 0)){
        print('fizzbuzz')
        }else if (i %%  3 == 0){
        print('fizz')
      }else if (i %% 5 == 0){
        print('buzz')
      }}

    vec_test <- function(start,stop,step){
      test_seq <- seq(start,stop,step)
      for (x in test_seq){
        res <- check_multiple(x)
        print(res)
      }}


    vec_test(0,50,5)
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.

    marioStairs <- function(s){
      steps <- (1:s)
      x <-"X"
      for (n in steps){
       print(noquote(strrep(x, n)))
      }
    }

    marioStairs(6)
    ```
